Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-07-13 unimodular-covers.tex
checking comments

\begin{definition}
We say that $P$ is a weak Minkowski summand of $Q$ if the normal fan of $Q$ refines that of $P$.
\end{definition}
The name is due to the fact that the condition on the refinement of the normal fans is equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$~\cite[??]{Grunbaum}. 
It has the following algebraic implication for the projective toric variety $X_Q$ corresponding to (the normal fan of) $Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the corresponding Cartier divisor $D_P$ on  $X_Q$ is \emph{numerically effective}, or ``nef''~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}; (observe that~\cite{CLS} use ``Minkowski summand'' meaning what we call ``weak Minkowski summand''). 
\paco{added this explanation}

\begin{definition}
The pair $(P, Q)$ has the integer decomposition property, or is IDP, if 
\begin{align*}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\end{definition}

In this language,  \Cref{conj:Oda}\eqref{itm:mixedIDP} can be rephrased as if $P$ is a weak Minkowski summand of $Q$ and $Q$ is smooth, then the pair $(P,Q)$ is IDP.

@@

This also implies the following two statements, also proved in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
\paco{Wait; ``width one'' is only a special case of ``prismatoid'' if we allow prismatoids to have vertices in two parallel \emph{faces}, not \emph{facets}. But \Cref{coro:prismatoid} is not so obvious in that more general case. Either prove it or remove the phrase ``special case''} \giulia{true...}
A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being $SL(\Z,d)$-equivalent to a Cayley sum.

@@

\begin{lemma}
\label{lemma:3<4}
Let $P$ be a lattice parallelepiped and let $T\subset P$ be an empty tetrahedron. Then, at least one of the four corner tetrahedra $T_i$ of the circumscribed parallelogram $C(T)$ is fully contained in $P$.
\end{lemma}

@@

Then $S=\conv(T\cup \{u\})$ can be triangulated in two different ways: $S=T \cup T'_4$, where $T'_4 = \conv(p_1, p_2, p_3, u) \subseteq T_4$ and $S= S_1 \cup S_2 \cup S_3$, with
\[
S_1= \conv(p_2,p_3,p_4, u),
S_2=\conv(p_1,p_3,p_4, u),
S_3=\conv(p_1,p_2,p_4, u).
\]
Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$, because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. By inductive hypothesis they can be covered unimodularly, and we are done,

@@

\begin{proof}[Proof of \Cref{thm:parallelepipeds}]\giulia{see comment above}
Arbitrarily triangulate the parallelepiped into lattice tetrahedra and apply \Cref{coro:coverpara} to these tetrahedra.
%Let $P \subseteq \R^3$ be a lattice parallelepiped. It can be triangulated into empty lattice tetrahedra, simply by taking a triangulation which uses all lattice points as vertices.  If all the tetrahedra in the triangulation are unimodular we are done. Otherwise, suppose $T$ is an empty tetrahedron in the triangulation which is not unimodular. \Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ circumscribed to $T$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point, which we denote by $u$. 
%
%
%This allows us to cover $T$ with lattice simplices of smaller volume recursively, until it is covered by unimodular ones.
\end{proof}

>>>>

\begin{definition}
We say that $P$ is a \emph{weak Minkowski summand} of $Q$ if the normal fan of $Q$ refines that of $P$.
\end{definition}
The name is due to the fact that the condition on the refinement of the normal fans is equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$~\cite[??]{Grunbaum}. 
It has the following algebraic implication for the projective toric variety $X_Q$ corresponding to (the normal fan of) $Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the corresponding Cartier divisor $D_P$ on  $X_Q$ is \emph{numerically effective}, or ``nef''~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}; (observe that~\cite{CLS} use ``Minkowski summand'' meaning what we call ``weak Minkowski summand''). 
\paco{added this explanation}

\begin{definition}
The pair $(P, Q)$ has the integer decomposition property, or \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\end{definition}

In this language,  \Cref{conj:Oda}\eqref{itm:mixedIDP} can be rephrased as: if $P$ is a weak Minkowski summand of $Q$ and $Q$ is smooth, then the pair $(P,Q)$ is IDP.

@@

This also implies the following two statements, also proved in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being $SL(\Z,d)$-equivalent to a Cayley sum.

@@

\begin{lemma}
\label{lemma:3<4}
Let $P$ be a lattice parallelepiped and let $T\subset P$ be a tetrahedron. Then, at least one of the four corner tetrahedra $T_i$ of the circumscribed parallelogram $C(T)$ is fully contained in $P$.
\end{lemma}

@@

Then $S=\conv(T\cup \{u\})$ can be triangulated in two different ways: $S=T \cup T'_4$, where $T'_4 = \conv(p_1, p_2, p_3, u) \subseteq T_4$ and $S= S_1 \cup S_2 \cup S_3$, with
\[
S_1= \conv(p_2,p_3,p_4, u),
S_2=\conv(p_1,p_3,p_4, u),
S_3=\conv(p_1,p_2,p_4, u).
\]
Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$, because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. The $S_i$ may not be empty, but we can triangulate them into empty tetrahedra, which by inductive hypothesis they can be covered unimodularly, and we are done.

@@

\begin{proof}[Proof of \Cref{thm:parallelepipeds}]\giulia{see comment above}
Arbitrarily triangulate the parallelepiped into empty lattice tetrahedra and apply \Cref{coro:coverpara} to these tetrahedra.
%Let $P \subseteq \R^3$ be a lattice parallelepiped. It can be triangulated into empty lattice tetrahedra, simply by taking a triangulation which uses all lattice points as vertices.  If all the tetrahedra in the triangulation are unimodular we are done. Otherwise, suppose $T$ is an empty tetrahedron in the triangulation which is not unimodular. \Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ circumscribed to $T$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point, which we denote by $u$. 
%
%
%This allows us to cover $T$ with lattice simplices of smaller volume recursively, until it is covered by unimodular ones.
\end{proof}
